---
title: "COVID 19"
author: "Juan Camilo Azuero"
date: "2/9/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this project is to explore how COVID has developed through time how it has affected different regions.

## Import libraries
```{r import_libraries, echo=T, results='hide',warning=FALSE,message=FALSE}
library("tidyverse")
library("zoo")
library("tidyr")
library("dplyr")
library("lemon")
library("lme4")
```


## Import data
Data is imported from a public repository of the Jhon Hopkins university 
```{r import_data, results='hide',warning=FALSE,message=FALSE}
base_url <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/"
access_type <- "?accessType=DOWNLOAD"
confirmed_filename <- "csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
deaths_filename <- "csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
recovered_filename <- "csse_covid_19_time_series/time_series_covid19_recovered_global.csv"
regions_filename <- "UID_ISO_FIPS_LookUp_Table.csv"
iso_codes_url <- "https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/raw/master/all/all.csv"

cases <- read_csv(paste(base_url,confirmed_filename,access_type,sep=""))
deaths <- read_csv(paste(base_url,deaths_filename,access_type,sep=""))
regions <- read_csv(paste(base_url,regions_filename,access_type,sep=""))
iso_codes <- read_csv(iso_codes_url)
```

It is important to note that the database is a compilation from multiple sources around the world, and that every country might have diferente measuring methodologies and effectiveness. This is an important source of bias that has to be accounted for.

## Initial transformations
```{r clean_data}
colnames(cases)[colnames(cases) == "Country/Region"] <- "Country"
colnames(deaths)[colnames(deaths) == "Country/Region"] <- "Country"
colnames(regions)[colnames(regions) == "Country_Region"] <- "Country"
# Delete info from places that are not countries, delete unnecesary columns
# and transform data into long format
cases <- cases %>% 
    filter(is.na(cases["Province/State"])) %>% 
    select(-c("Province/State","Lat","Long")) %>%
    gather("day","cases",-c("Country"))
deaths <- deaths %>% 
    filter(is.na(deaths["Province/State"])) %>% 
    select(-c("Province/State","Lat","Long")) %>%
    gather("day","deaths",-c("Country"))
regions <- regions %>% filter(is.na(regions["Province_State"])) %>% select("Country","iso3","Population")
iso_codes <- iso_codes %>% select("alpha-3","region")
colnames(iso_codes)[colnames(iso_codes) == "alpha-3"] <- "iso3"
regions <- merge(regions, iso_codes, by="iso3")
# Merge datasets into summary
summary <- merge(cases,deaths,by=c("Country","day"))
summary <- merge(summary,regions,by="Country", all.y = FALSE)
summary <- summary %>% filter(!is.na(summary['Population']))

summary$day <- as.Date(summary$day, format = "%m/%d/%y")
summary <- summary[order(summary["Country"], summary["day"]),]
# Obtains daily info from cumulative columns
summary_aux <- summary %>% select("Country","day","cases","deaths") %>% mutate(day = day+1)
colnames(summary_aux)[colnames(summary_aux) == "cases"] <- "cases_lag"
colnames(summary_aux)[colnames(summary_aux) == "deaths"] <- "deaths_lag"
summary <- merge(summary,summary_aux,by=c("Country","day"), all.y=FALSE)
summary['cases_day'] = summary['cases']-summary['cases_lag']
summary['deaths_day'] = summary['deaths']-summary['deaths_lag']
summary <- summary %>% select(-c("cases_lag","deaths_lag"))
# Normalize per million
summary['cases_per_million'] = 1000000*summary['cases']/summary['Population']
summary['deaths_per_million'] = 1000000*summary['deaths']/summary['Population']
summary['cases_day_per_million'] = 1000000*summary['cases_day']/summary['Population']
summary['deaths_day_per_million'] = 1000000*summary['deaths_day']/summary['Population']
# Global summary
global_summary <- summary %>% group_by(day) %>% summarize(cases=sum(cases), deaths=sum(deaths),
                                                         cases_day=sum(cases_day), deaths_day=sum(deaths_day))
global_summary["cases_day_wa"] = rollmean(global_summary["cases_day"],7,c(NA,NULL,NA))
global_summary["deaths_day_wa"] = rollmean(global_summary["deaths_day"],7,c(NA,NULL,NA))
global_summary["mortality_rate"] <- global_summary["deaths"]/global_summary["cases"]
global_summary_region <- summary %>% group_by(day, region) %>% summarize(cases=sum(cases), deaths=sum(deaths),
                                                         cases_day=sum(cases_day), deaths_day=sum(deaths_day),
                                                         population=sum(Population))
global_summary_region['cases_day_per_million'] = 1000000*global_summary_region['cases_day']/global_summary_region['population']
global_summary_region['deaths_day_per_million'] = 1000000*global_summary_region['deaths_day']/global_summary_region['population']
```
## Global analysis
First let's plot the evolution over time of the total number cases and deaths

```{r plots_1, fig.width=7, fig.height=4, warning=FALSE}
coeff <- 100
deaths_color <- rgb(0.9, 0.2, 0.1, 1)
cases_color <- rgb(0.2, 0.6, 0.9, 1)
global_summary %>% select("day","cases","deaths") %>% 
  ggplot(aes(x=day)) + 
  geom_line(aes(y=cases),color=cases_color) + 
  geom_line(aes(y=deaths*coeff),color=deaths_color) + 
  xlab("Date") + 
  scale_y_continuous(
    name = "Cases",
    sec.axis = sec_axis(~./coeff, name="Deaths")
  ) +
  theme(
    axis.title.y = element_text(color = cases_color),
    axis.title.y.right = element_text(color = deaths_color),
    plot.title = element_text(size=14, face="bold.italic")
  ) + 
  ggtitle("Cases and deaths")
```

Now let's repeat the previous plot, but using a logarithmic axis for the amount of cases and deaths.

```{r plots_2, fig.width=7, fig.height=4, warning=FALSE}
global_summary %>% select("day","cases","deaths") %>% 
  ggplot(aes(x=day)) + 
  geom_line(aes(y=cases,colour="Cases")) + 
  geom_line(aes(y=deaths,colour="Deaths")) +
  xlab("Date") + 
  ylab("Number of people") + 
  scale_y_continuous(trans='log10') +
  scale_colour_manual("", breaks = c("Cases", "Deaths"), values = c(cases_color,deaths_color)) + 
  ggtitle("Cases and deaths")+
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

The first plot shows how cases and deaths has risen consistently since the start of 2020. The plot also shows how since 2022 the number of cases per day increased considerably due to the omicron variant.

As the nature of contagion is exponential, the second graph is very useful to understand the evolution of the pandemic. The logarithmic transformation serves to see how high exponential growth in the first 3 months decayed until today. The wave because of omicron at the beginning of 2022 produces an upward turn of the cases, but deaths stayed flat.

In order to further explore the evolution of cases and deaths through time, lets plot also the rate of change for these 2 variables.

```{r plots_3, fig.width=7, fig.height=4, warning=FALSE}
global_summary %>% select("day","cases","deaths","cases_day_wa","deaths_day_wa") %>% 
  ggplot(aes(x=day),yaxp = c(0, 1, 10)) + 
  geom_line(aes(y=100*cases_day_wa/cases,colour="Cases")) + 
  geom_line(aes(y=100*deaths_day_wa/deaths,colour="Deaths")) +
  xlab("Date") + 
  ylab("Percentage increase") + 
  scale_y_continuous(trans='log10',breaks=c(0,0.1,0.2,0.5,1,2,5,10,20,50,100)) +
  scale_colour_manual("", breaks = c("Cases", "Deaths"), values = c(cases_color,deaths_color)) + 
  ggtitle("Rate of change for deaths and confirmed cases") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

This plot displays how cases and deaths increased daily by approximate 30% at the beginning of the pandemic, and has decayed until reaching levels of daily change of 0.2% and 0.1% for cases and deaths respectively.

Now let's plot the number of daily cases and deaths. 

```{r plots_4, fig.width=7, fig.height=4, warning=FALSE}
global_summary %>% select("day","cases_day_wa") %>% 
  ggplot(aes(x=day, y=cases_day_wa)) + 
  geom_line() + 
  xlab("day")+ 
  ylab("Daily infections") +
  ggtitle("Weekly average of cases per day") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
global_summary %>% select("day","deaths_day_wa") %>% 
  ggplot(aes(x=day, y=deaths_day_wa)) + 
  geom_line() + 
  xlab("day") +
  ylab("Daily deaths") +
  ggtitle("Weekly average of deaths per day") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

In the plot of daily cases a steady increase in all 2020 is shown ending in a peak of cases at the end of the year. During 2021 there were another 2 similar peaks, and then in December the number of cases skyrocketed because of the omicron variant. The plot of daily deaths has some considerable differences including an early peak in April and the absence of immense growth in the end of 2021. The early peak is probably not seen in the cases plot because detection was not as good at the beginning of the pandemic, and there is no mortality increase in the end of 2021 thanks to the rollout of the vaccine and the lower severity of omicron infections. The following plot displays how the measured mortality peaks in April (probably because of lack of diagnostic tools) stabilizes at 2% in 2021 and turn downward in 2022 for the reasons mentioned.

```{r plots_5, fig.width=7, fig.height=4, warning=FALSE}
global_summary %>% select("day","mortality_rate") %>% 
  ggplot(aes(x=day, y=mortality_rate)) + 
  geom_line() + 
  xlab("day") +
  ylab("Mortality rate (deaths/cases)") +
  ggtitle("Acumulated mortality") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

## Analysis by region
The next two plots show how the pandemic evolved in each continent. The Americas and Europe were the continents that took the most damage as they display more infections and deaths than the other continents. The number of deaths in Europe clearly display 2 valleys while the graph for the Americas is considerably flatter.

```{r plots_region, fig.width=7, fig.height=4, warning=FALSE}
global_summary_region %>% select("day","cases_day_per_million","region") %>% 
  ggplot(aes(x=day, y=cases_day_per_million,colour=region)) + 
  geom_line() + 
  xlab("day")+ 
  ylab("Daily infections") +
  ggtitle("Infections per million people") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
global_summary_region %>% select("day","deaths_day_per_million","region") %>% 
  ggplot(aes(x=day, y=deaths_day_per_million,colour=region)) + 
  geom_line() + 
  xlab("day") +
  ylab("Daily deaths") +
  ggtitle("Deaths per million people") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```


## Analysis by country
Now let's analyze how the outbreak of COVID affected the different countries. Seeking to avoid outliers produced by small countries it was decided to filter all countries with less than 1 million habitants.
```{r country_etl}
summary_total <- summary %>% group_by(Country) %>% summarize(
  cases=sum(cases_day),
  deaths=sum(deaths_day),
  population=mean(Population),
  region=first(region)
)
summary_total["mortality_rate"] <- summary_total["deaths"]/summary_total["cases"]
summary_total["deaths_per_million"] <- 1000000*summary_total["deaths"]/summary_total["population"]
summary_total["cases_per_million"] <- 1000000*summary_total["cases"]/summary_total["population"]
summary_total_big <- summary_total %>% filter(population > 1000000)
```

Let's organize countries by the normalized amount of deaths, cases and by the mortality rate to see what we can find.

### Least deaths per million people
```{r tabla_1, render=lemon_print}
head(summary_total_big[order(summary_total_big$deaths_per_million),],15)
```

### Most deaths per million people
```{r tabla_2,render=lemon_print}
tail(summary_total_big[order(summary_total_big$deaths_per_million),],15)
```

The countries with higher mortality rates are principally countries from eastern Europe where omicron has been particularly severe, Peru where the health system collapsed catastrophically in the first wave, and countries like Brazil and the USA where distancing and other measures were famously dismissed by their presidents.

On the other hand, very surprisingly the countries with the least number of deaths per million people are African countries with limited exceptions like New Zeland and Tanzania. In  [this](https://www.bbc.com/news/world-africa-54418613) BBC report Anne Soy points to the early adoption of distancing and other measures, the support from the public, the relatively young population, the warmer climate and the good health systems which had previous experience dealing with viruses like Ebola.

Now let's explore the number of reported infections 

### Least reported infections per million people
```{r tabla_3,render=lemon_print}
head(summary_total_big[order(summary_total_big$cases_per_million),],15)
```

### Most reported infections per million people
```{r tabla_4,render=lemon_print}
tail(summary_total_big[order(summary_total_big$cases_per_million),],15)
```

The countries with the least number of reported infections per million are African countries for the reasons commented before and also probably influenced by under reporting. The countries with most number of cases per million are countries from eastern Europe that also have relatively large amounts of deaths, and countries with low deaths per million that did a very good job in testing and isolating infected people.

The two tables below display the countries with higher and lower mortality rates.

### Lower mortality rates
```{r tabla_5,render=lemon_print}
head(summary_total_big[order(summary_total_big$mortality_rate),],15)
```

### Higher mortality rates
```{r tabla_6,render=lemon_print}
tail(summary_total_big[order(summary_total_big$mortality_rate),],15)
```

Assuming (probably incorrectly but not to far from reality) that the mortality rate of the virus is homogeneous trough the world, it can be concluded that the reported mortality rates correlate negatively with the effectiveness of the countries testing campaigns, and that the lower mortality rates are close to the virus real mean. The tables above then show how developed countries principaly from Europe and Asia excelled regarding their testing campaigns while countries from Africa, Latin America and the Middle East were not able to diagnosticate their citizens effectively. It can also be concluded that the real mortality of the virus is close to 0.15%.

Finally let's visualize a scatter plot of the deaths vs number of infections per million.

```{r plots_6, fig.width=7, fig.height=4, warning=FALSE}
summary_total_big %>% select("deaths_per_million","mortality_rate","Country","population","region") %>% 
  ggplot(aes(x=mortality_rate, y=deaths_per_million,label=Country,color=region)) + 
  geom_text(
    aes(label=ifelse(deaths_per_million>3000|mortality_rate>0.035,as.character(Country),'')),
    size=3,
    hjust=-0.1, 
    vjust=-0.1
  ) +
  geom_point(aes(size=population)) + 
  xlab("Mortality rate (deaths/cases)") +
  ylab("Deaths per million") +
  xlim(0, 0.1) +
  ggtitle("Cases vs Deaths per million") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

The countries that appear lower on the y axis have the lowest number of deaths and represent places where the pandemic had small impact. while the x axis correlates with the inversely with the effectivity of the diagnostic effort of the country. This plot shows a cluster of African countries with high mortality rates and low number of deaths per million and a cluster of eastern European countries with high death toll as discussed before.

### Correlation between mortality rates and deaths
Now let's use a lineal model to model the relation between the mortality rate and the number of deaths per million.
```{r model}
lin_reg <- lm(deaths_per_million ~ mortality_rate,data = summary_total_big)
summary(lin_reg)
```

This summary shows that the coefficient for mortality rate estimated is not significative, and as seen in the plot below the linear fit is not very good.

```{r prediction, fig.width=7, fig.height=4, warning=FALSE}
summary_total_big <- summary_total_big %>% mutate(pred=predict(lin_reg))
summary_total_big %>% select("deaths_per_million","mortality_rate","Country","pred", "region") %>% 
  ggplot(aes(x=mortality_rate)) + 
  geom_point(aes(y=deaths_per_million,colour=region)) +
  geom_point(aes(y=pred),shape=3) + 
  xlab("Mortality rate (deaths/cases)") +
  ylab("Deaths per million") +
  xlim(0, 0.1) +
  ggtitle("Cases vs Deaths per million") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

As seen in the previous differents regions of the world seem to have different relations between the mortality rate and the number of deaths per million people. In order to consider this effect in the analysis let's estimate a mixed linear model. By using the lme4 library we can estimate an intercept and coefficient for every region. For more infomation regarding mixed linear models you can refer to the [documentation](https://rdrr.io/cran/lme4/man/lmer.html) of lme4.

```{r mixed_linear_model, warning=FALSE}
mixed_model <- lmer(deaths_per_million ~  1 + mortality_rate + (1 + mortality_rate|region),summary_total_big)
summary_total_big["prediction"] = mixed_model %>% predict(summary_total_big)
summary(mixed_model)
```

The mixed linear model estimated is based on the asumption that the intercept and coeficient of the mortality rate is composed by a gobal fixed effect and a random effect that has a normal distribution centered at 0. The model also asumes that the residuals are normally distributed and independent of the predicted values. Based on this, lme4 finds a standard error and a t statistic asociated with the fixed effects. The summary of the model show that the intercept is positive and significant at a 95% level, while the coefficient of the mortality rate in only significant at the 80% level. The positive coefficient of the mortality rate indicates that higher mortality rates (posible product of bad diagnostic campaigns) relates to a higher deaths in the country.

The next two plots show that the residuals has relatively low heteroscedasticity and distribute aproximately normal. 

```{r mixed_linear_model_residuals, fig.width=7, fig.height=6, warning=FALSE}
plot(fitted(mixed_model), resid(mixed_model))
qqnorm(resid(mixed_model))
qqline(resid(mixed_model), col="red")
```

Now let's visualize the resulting regresions for every region.

```{r mixed_linear_model_prediction, fig.width=7, fig.height=6, warning=FALSE}
summary_total_big %>% select("deaths_per_million","mortality_rate","Country","prediction","region") %>%
  ggplot(aes(x=mortality_rate)) +
  geom_point(aes(y=deaths_per_million,colour=region)) +
  geom_line(aes(y=prediction,colour=region)) +
  xlab("Mortality rate (deaths/cases)") +
  ylab("Deaths per million") +
  xlim(0, 0.1) +
  ggtitle("Cases vs Deaths per million") +
  theme(plot.title = element_text(size=14, face="bold.italic"))
```

The plot before shows a much beter fit than the model that did not consider the region of the country, but it still can be improved. This analysis hints of a positive correlation between mortality rate and deaths per million in Europe and the Americas, while in Africa, Asia and Oceania no correlation was captured. This does not imply that there is no relation between the two variables, including some predictive unused attribute might reduce the error and reveal a stronger relation between the pandemic statistics analyzed. So in order to deepen this analysis one might add other variables like economic output, mobility trends and a more specific region. Another possible next steps are consider the time dimension,  handle outliers like Syria and Afghanistan where war raged during the pandemic, or include that information in the model.

#### Session info
```{r session_info}
sessionInfo()
```